The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum

### 2013-12-05 sections/climate.tex
Address minor editor comments

The spatial distributions of JJA air surface temperatures and DJF precipitation rates from the four reanalysis climatologies are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}. Summer temperature and winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation. Because reanalyses include observational information through data assimilation, they are closer to observations than pure GCMs in densely monitored regions, such as the southern part of our modelling domain \citep{bengtsson-etal-2007}. On the other hand, the atmospheric physics embedded in the models provide more realistic output than obtained by spatial interpolation in sparsely monitored regions, such as the northern part of our modelling domain.

@@

WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the 10\,km-resolution model grid using the module r.proj from GRASS~GIS \citep{soft:grass}. In addition, WorldClim data was extrapolated to cover grid points in the ocean using the nearest-neighbour algorithm \citep{soft:scipy}. Note that here we choose to present original rather than interpolated data, in order to highlight differences between datasets related to spatial resolution (Figs.~\ref{fig:temp} and~\ref{fig:prec}).

The CFSR climatology presents wave-like artefacts in its precipitation field (Fig.~\ref{fig:prec}). These are a common feature in GCM simulations due to the failure of spectral models to resolve the very local nature of topography-induced precipitation events, which becomes particularly visible in the high-resolution, less spatially smoothed data. Therefore, an alternative forcing was prepared for CFSR by smoothing its precipitation field. This was achieved by averaging data locally in a circular kernel of 7\,pixels in diameter prior to re-projection using the module r.neighbors from GRASS~GIS \citep{soft:grass}. Finally, input files for PISM were prepared using \citet{web:nc4py}.

>>>>

The spatial distributions of JJA air surface temperatures and DJF precipitation rates from the four reanalysis climatologies are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}. Summer temperature and winter precipitation are most relevant to the glacier model as they drive summer melt and winter accumulation, respectively. Because reanalyses include observational information through data assimilation, they are closer to observations than pure GCMs in densely monitored regions, such as the southern part of our modelling domain \citep{bengtsson-etal-2007}. On the other hand, the atmospheric physics embedded in the models provide more realistic output than obtained by spatial interpolation in sparsely monitored regions, such as the northern part of our modelling domain.

@@

WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to the Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the 10\,km-resolution model grid using the module r.proj from GRASS~GIS \citep{soft:grass}. In addition, WorldClim data was extrapolated to cover grid points in the ocean using the nearest-neighbour algorithm \citep{soft:scipy}. Note that here we choose to present original rather than interpolated data, in order to highlight differences between datasets related to spatial resolution (Figs.~\ref{fig:temp} and~\ref{fig:prec}).

The CFSR climatology presents wave-like artefacts in its precipitation field (Fig.~\ref{fig:prec}). These are a common feature in GCM simulations due to the failure of spectral models to resolve the very local nature of topography-induced precipitation events, which becomes particularly visible in the high-resolution, less spatially-smoothed data. Therefore, an alternative forcing was prepared for CFSR by smoothing its precipitation field. This was achieved by averaging data locally in a circular kernel of 7\,pixels in diameter prior to re-projection using the module r.neighbors from GRASS~GIS \citep{soft:grass}. Finally, input files for PISM were prepared using \citet{web:nc4py}.

### 2013-12-05 sections/discuss.tex
Address minor editor comments

Most likely, this is the signature of an orographic precipitation effect. As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid. However the ability of a GCM to reproduce these contrasts in precipitation is bounded by spatial resolution. In a model of coarser resolution, these high mountain ranges are reduced to smooth hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}). However this effect alone cannot explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Figs.~\ref{fig:precheatmap}--\ref{fig:precdiff}) despite of its high spatial resolution (Table~\ref{tab:reanalyses}).

@@

To distinguish the effects of temperature and precipitation biases on the ice sheet model, we run a series of eight additional simulations. These new simulations use a “hybrid” climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa. A single temperature offset of 5\,\unit{\degree C} is used. This value was chosen to allow comparison with previous results (Fig.~\ref{fig:cool05}). In each case, temperature lapse rates are computed using the reference topography from the corresponding dataset.

@@

We compare numerically modelled ice sheet geometries to a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating (Fig.~\ref{fig:locmap}). This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than for the Laurentide ice sheet \citep{porter-swanson-1998,dyke-2004,stroeven-etal-2010,stroeven-etal-inpress}.

>>>>

Most likely, this is the signature of an orographic precipitation effect. As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid. However the ability of a GCM to reproduce these contrasts in precipitation is bounded by spatial resolution. In a model of coarser resolution, these high mountain ranges are reduced to smooth hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}). However, this effect alone cannot explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Figs.~\ref{fig:precheatmap}--\ref{fig:precdiff}) despite of its high spatial resolution (Table~\ref{tab:reanalyses}).

@@

To distinguish the effects of temperature and precipitation biases on the ice sheet model, we run a series of eight additional simulations. These new simulations use a “hybrid” climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa. A single temperature offset of 5\,\unit{\degree C} is used. This value was chosen to allow for comparison with previous results (Fig.~\ref{fig:cool05}). In each case, temperature lapse rates are computed using the reference topography from the corresponding dataset.

@@

We compare modelled ice sheet geometries to a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating (Fig.~\ref{fig:locmap}). This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than for the Laurentide ice sheet \citep{porter-swanson-1998,dyke-2004,stroeven-etal-2010,stroeven-etal-inpress}.

### 2013-12-05 sections/intro.tex
Address minor editor comments

The other side of the coin is that avoiding this circular dependence unfortunately goes hand-in-hand with a need for simplifying assumptions regarding Earth's past climate. Such studies include energy balance modelling approaches \citep{tarasov-peltier-1997} and geographic parametrizations of surface mass balance \citep{robert-1991} or climate forcing \citep{johnson-fastook-2002}. Standing on a middle ground, temperature offset methods \citep{greve-etal-1999,bintanja-etal-2005} make use of the high level of detail available in present climate datasets such as gridded observation datasets, GCM output or reanalyses, while using simplifying representations of past climate deviations from the present state.

Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating the responses of a numerical model, in terms of glacier extent, to inputs from several climate datasets. This is not entirely a new approach as a few studies of this kind are presently available. \citet{quiquet-etal-2012}, for example, assessed the sensitivity of a Greenland ice sheet model to various climate forcing fields, including a regional parametrization by \citet{fausto-etal-2009}, output from several GCMs and an atmospheric reanalysis. \citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of Northern Hemisphere ice sheets to climate forcing from different PMIP LGM simulations. Although using different set-ups, all three studies demonstrate the very large sensitivity of ice sheet models to the choice of climate forcing data used. In our study, to limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data. Hence, rather than using GCM output, we force our model with climate reanalysis data, which builds on observational information through data assimilation \citep{bengtsson-etal-2007}. Furthermore, we focus our study regionally on the former Cordilleran Ice Sheet in western North America.

@@

Here we use PISM, a Parallel Ice Sheet Model \citep{web:pism}, to simulate the extent and thickness of the Cordilleran Ice Sheet at the LGM (Fig.~\ref{fig:locmap}). We force our model with multiple climate datasets and compare the modelled ice extent to a geomorphological reconstruction of the LGM ice sheet margin by \citet{dyke-2004}. Thereby, we aim to determine the climate dataset which is most suited for simulation of the Cordilleran Ice Sheet and to use as input for future, transient studies over a glacial cycle. To our knowledge, this is the first modelling study that specifically focuses on the Cordilleran Ice Sheet since \citet{robert-1991}.

>>>>

However, avoiding this circular dependence requires simplifying assumptions regarding Earth's past climate. Such studies include energy balance modelling approaches \citep{tarasov-peltier-1997} and geographic parametrizations of surface mass balance \citep{robert-1991} or climate forcing \citep{johnson-fastook-2002}. Alternatively, temperature offset methods \citep{greve-etal-1999,bintanja-etal-2005} make use of the high level of detail available in present climate datasets such as gridded observation datasets, GCM output or reanalyses, while using simplifying representations of past climate deviations from the present state.

Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating the responses of a numerical model, in terms of glacier extent, to inputs from several climate datasets. This is not entirely a new approach as a few studies of this kind are presently available. \citet{quiquet-etal-2012}, for example, assessed the sensitivity of a Greenland ice sheet model to various climate forcing fields, including a regional parametrization by \citet{fausto-etal-2009}, output from several GCMs and an atmospheric reanalysis. \citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of Northern Hemisphere ice sheets to climate forcing from different PMIP LGM simulations. Although using different set-ups, all three studies demonstrate the very large sensitivity of ice sheet models to the choice of climate forcing data used. To limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data. Hence, rather than using GCM output, we force our model with climate reanalysis data, which builds on observational information through data assimilation \citep{bengtsson-etal-2007}. Furthermore, we focus our study regionally on the former Cordilleran Ice Sheet in western North America.

@@

Here we use PISM, a Parallel Ice Sheet Model \citep{web:pism}, to simulate the extent and thickness of the Cordilleran Ice Sheet at the LGM (Fig.~\ref{fig:locmap}). We force our model with multiple climate datasets and compare the modelled ice extent to a geomorphological reconstruction of the LGM ice sheet margin by \citet{dyke-2004}. We aim to determine the climate dataset which is most suited for simulation of the Cordilleran Ice Sheet and to use as input for future, transient studies over a glacial cycle.
